---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

```{r}

fileName <- "activity.zip"
fileNameCSV <- gsub("zip", "csv", fileName)

unzip(fileName, fileNameCSV)
data <- read.csv(fileNameCSV, header=TRUE, sep=",", na.strings="NA")
summary(data)
```

## What is mean total number of steps taken per day?

Get the total number of steps take per day

```{r}
totalSteps <- tapply(data$steps, data$date, FUN=sum, na.rm =TRUE)
```

Make a histogram of the total number of steps taken each day

```{r}
hist(totalSteps, xlab="total steps per day", ylab="frequency", main="histogram of total steps per day")
```

Calculate the mean and median of the total number of steps taken per day

```{r}
mean(totalSteps, na.rm=TRUE)
median(totalSteps, na.rm=TRUE)
```

## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
meanSteps <- tapply(data$steps, data$interval, FUN=mean, na.rm = TRUE)
plot(meanSteps, type="l", 
     xlab="5-min time intervals", 
     ylab="mean # of steps taken", 
     main="average steps taken at 5-min time intervals"
)
```

Get 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps?

```{r}
positionOfMax <- which.max(meanSteps)
names(positionOfMax)
```

## Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset

```{r}
sum(is.na(data$steps))
```

2. As a strategy for filling in all of the missing values in the dataset we use the mean for that 5-minute interval.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
for(i in 1:length(data$steps)){
  if(is.na(data[i,1])){
    data[i,1] <- unname(meanSteps[toString(data[i,3])])
  }
}
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
totalSteps <- tapply(data$steps, data$date, FUN=sum, na.rm =TRUE)
hist(totalSteps, xlab="total steps per day", ylab="frequency", main="histogram of total steps per day")
mean(totalSteps, na.rm=TRUE)
median(totalSteps, na.rm=TRUE)
```

Values are higher than thoes calculate in the first part of the assighnment since the NA values tend to shift the results towards lower values.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
for(i in 1:length(data[,1])){
  name = weekdays(as.Date(data$date[i]))
  if (name == "Saturday" | name == "Sunday"){
    data$category[i] <- "weekend"
  } else {
    data$category[i] <- "weekday"
  }
}
data$category <- as.factor(data$category)
```

2. Make a panel plot containing a time series plot  of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
library(ggplot2)
library(plyr)
tmpdata <- ddply(data, .(interval,category), summarize, steps = mean(steps, na.rm=TRUE))
ggplot(data=tmpdata, aes(x=interval, y=steps, group=category)) + geom_line()+ facet_wrap(~ category, nrow=2)
```

In the moring hours until around 8am there is more activity on the weekdays, whereas in the afternoon there is more activity on the weekend. 